Making multiple-locus association mapping on a genome-wide scale routine

### 2017-10-18 AM+draft.tex
181017 Working on methods

The Eagle approach for multiple-locus association mapping is based on building a linear mixed model, iteratively, via forward 
selection.  Suppose $s$ iterations of our model building process have already been performed. This means $s$ snp-trait 
associations which identify $s$ separate genomic regions of interest have been found.  To perform the $s+1$ iteration of 
our model building procedure, we perform the following.

First, we fit the current model to the data. 
Let $y^{(n \times 1)}$ be the vector containing the quantitative trait data. 
Let $M^{(n \times L)} = [m_1 m_2 \ldots m_L]$ be the matrix containing the genotype data which has been collected 
from $L$ loci that span the genome. Here, the vector $m_j^{(n \times 1)}$ contains the $j$th snp genotypes for the $n$ individuals. 
The genotypes are coded as -1, 0, and 1 corresponding to snp genotypes AA, AB, and BB, respectively.

The model is of the form 
\begin{equation}
y = X \tau + u_g + e
\end{equation}
where 
$X^{(n \times p)}$ is a known design matrix of full rank and may contain data on non-genetic effects such as age and height 
or principal component values to account for population structure (ref). The vector 
$\tau^{(p \times 1)}$ has $p$ fixed effects parameters. The vector 
$u_g^{(n \times 1)}$ contains the 
genetic effects. The vector of residuals is 
$e^{(n \times 1)}$ whose distribution is assumed to follow $N(0, \sigma^2_e I^{(n \times n)})$.

Up to this point,  this model differs little from standard linear mixed models for association mapping (refs). 
However, 
it is how we specify $u_g$ that distinguishes our model from others. The genetic effects $u_g$ are modelled as

Suppose $s$ iterations have been performed of our multiple locus association mapping algorithm. This means $s$ snp-trait 
associations of significance have been identified. It also means $s$ separate regions containing putative genes 
that influence the trait have been mapped.  To find perform iteration $s+1$ and the next snp, we perform the following steps.  First,

Let $M^{(n \times L} $be a matrix of snp genotypes which have been collected from the n individuals across L snp loci. These L loci are assumed to span the entire genome. M contains values -1, 0, and 1 corresponding to snp genotypes aa, ab, and bb, respectively.

S={S1,...Ss} be

Our model building process is as follows.  Suppose $s$ SNP have already been found. First we fit the current model to the data.

Suppose our iterative model building procedure has already found $s$ snp. Then

Our linear mixed model for multiple-locus association mapping, given trait data $y^{(n \times 1)}$ which are collected on $n$ individuals, 
is of the form 
\begin{equation}
y = X \tau + u_g + e
\end{equation}
where $X^{(n \times p)}$ is a known design matrix of full rank, $\tau^{(p \times 1)}$ is a vector of fixed effects parameters, $u_g^{(n \times 1)}$ is a vector 
of genetic effects, and $e^{(n \times 1)}$ is a vector of residuals whose distribution is assumed to follow $N(0, \sigma^2_e I^{(n \times n)})$.

Now, let $M^{(n \times L)}$ be a matrix of genotypes collected from $L$ snp.  The 
matrix contains the values -1, 0, and 1, corresponding to the snp genotypes AA, AB, and BB, respectively.  It is assumed that 
missing genotypes have been imputed.

Suppose $s$ SNP have already been found.

$S={ S_1, S_2, \ldots, S_s}$ be the set of column indexes

\bibliographystyle{nature}
\bibliography{bibcitelibrary}

>>>>

The Eagle approach for multiple-locus association mapping is based on building a linear mixed model, iteratively, via forward 
selection.  Suppose $s$ iterations of our model building process have already been performed. This means $s$ snp-trait 
associations have been identified.  It also means that $s$ separate genomic regions of interest have been found.

To perform the $s+1$th  iteration of 
our model building procedure, we do the following. 
First, we fit the current model to the data. 
Let $S = \{ S_1, S_2, \ldots, S_s\}$ be a set of ordinal numbers where $S_k$ is the $S_k$th ordered snp that was 
selected in the $k$th iteration. For example, if three iterations of our model building procedure 
have been performed and say the 500023rd, 15th, and 420th, 
snp were selected, then $S=\{500023, 15, 420\}$. 
Let $y^{(n \; \times \;1)}$ be a vector containing the quantitative trait data. 
Let $M^{(n \; \times \; L)} = [m_1 m_2 \ldots m_L]$ be a matrix containing the genotype data which have been collected 
from $L$ loci that span the genome.  The columns of this matrix are most commonly placed in map ordered but 
this is not a requirement. 
The vector $m_j^{(n \; \times \; 1)}$ contains the genotypes, from the $n$ individuals,  for the $j$th snp. 
The genotypes are coded as -1, 0, and 1 corresponding to snp genotypes AA, AB, and BB, respectively.

gocard123

The (current) model is of the form 
\begin{equation}
\label{eq1}
y = X \tau + u_g + e
\end{equation}
where 
$X^{(n \; \times \; p)}$ is a known design matrix of full rank and may contain data on non-genetic effects such as age and height 
and/or principal component values to account for population structure (ref). The vector 
$\tau^{(p \; \times \; 1)}$ has $p$ fixed effects parameters. The vector 
$u_g^{(n \; \times \; 1)}$ contains the 
genetic effects. The vector of residuals is 
$e^{(n \; \times \;1)}$ whose distribution is assumed to follow $N(0, \sigma^2_e I^{(n \; \times \; n)})$. 
So far,  this model differs little from standard linear mixed models for association mapping (refs). 
However, 
it is how we specify $u_g$ that distinguishes our model from others.

The genetic effects $u_g$ are modelled as 
\begin{equation}
\label{eq2}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + M_{-S} a_{-S}
\end{equation}
where $m_{S_k}^{(n \; \times \; 1)}$ is the vector of genotypes for the $S_k$th snp locus which is the $k$th selected snp, 
$a_{S_k}$ is the additive effect of the $S_k$th snp locus, $M_{-S}^{(n \; \times \; L-s)}$ is the matrix of  snp genotypes 
with the data for the selected snp in $S$ removed,  and $a_{-S}^{(L-s \; \times  \; 1)}$ is a random effect whose distribution is 
$a_{-S} \sim N(0, \sigma_a^2 I^{(L-s \; \times \;  L-s)})$. 
The sum of terms on the left hand side are fixed effects.  The other term is a random effect.  The sum of terms accounts 
for the additive effects of genes that are in linkage disequilibrium with the selected snp. The other term models 
snp-trait associations along the entire genome, except for those snp that have already been selected. This is a very simple genetic model but it 
is effective for discovering snp-trait associations. It also reduces the computational cost of our model building procedure.

%The simplicity of our genetic model also reduces the cost of computation.

Second, we estimate the parameters of (\ref{eq1}) and (\ref{eq2}) via residual maximum likelihood estimation.

Third, we identify the $(s+1)$th snp that is in strongest association with the trait, based on the maximum score statistic
$t_j^2 = \frac{ \widetilde{a} _j^2}{\textrm{var}(\widetilde{a}_j)}$ where $\widetilde{a}_j$ is the best linear unbiased predictor (BLUP), 
and $\textrm{var}(\widetilde{a}_j)$ is its variance. This statistic is not only appealing intuitively, where we 
identify a snp based on its (random) effect size and accuracy, but is theoretically justified.  It follows from outlier detection 
in linear and linear mixed models (ref).

Fourth, we determine the importance of the $(s+1)$th selected snp via a model selection strategy. 
We begin by reforming (\ref{eq2}) where $S$ now contains the $s + 1$ selected snp.  We then fit this new model to the data
via maximum likelihood and calculate its extended Bayesian information criteria \cite{chen2008extended}.

We reform (\ref{eq2}) where 
$S$ contains $s+1$ selected snp. We then fit (\ref{eq1}) and (\ref{eq2}) to the data via maximum likelihood and calculate $extBIC$.
We found this model selection criterium to perform better for association mapping problems than standard model selection 
criteria such as BIC and IAC, as was also found by XXXXX.  If the extBIC value is less than the extBIC for the previous model, 
then we add the $s+1$th snp  to the model and repeat the process. If  the extBIC value is greater than or equal to the extBIC for the 
previous model, then our iterative procedure is finished.  The set of snp in strongest association with the trait is the $s$ snp
previously identified.

\bibliographystyle{naturemag}
\bibliography{bibcitelibrary}
